---
title: "Assignment 2: Coding Basics"
author: Ricky Prophete
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1.  
seq(1, 100, 4) #create sequence function
one_hundred_by_four <- seq(1, 100, 4) #assign name to sequence function

#2. 
mean(one_hundred_by_four) #compute mean of sequence function
median(one_hundred_by_four) #compute median of sequence function 

#3. 
x <- mean(one_hundred_by_four) #set x == to mean of sequence function
y <- median(one_hundred_by_four) #set y == to median of sequence function 

x > y 
```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
#a.
vector_student_names <- c("Sally", "Jack", "Chris", "Miranda") #character vector
#b.
vector_test_score <- c(58, 46, 83, 94) #numeric vector
#c.
vector_pass_result <- c(TRUE, FALSE, TRUE, TRUE) #logical vector 
#data frame
dataframe_student_performance <- data.frame(vector_student_names,vector_test_score, vector_pass_result)
names(dataframe_student_performance) <- c("Name","Score","Passed"); View(dataframe_student_performance) 

dataframe_student_performance #dataframe test

```

9. QUESTION: How is this data frame different from a matrix?

> Answer: This data frame differs in that it combines elements of different types (numeric, logical, & character), whereas a matrix can only contain elements of the same type.

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
#10.
logical_test <- function(x){
  ifelse(x >= 50, print(TRUE), print(FALSE))  
} #determine whether score is >= 50

#11.
logical_test(vector_test_score) #apply function to scores
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 'ifelse' worked as a shorthand alternative to the inclusion of both 'if' and 'else'. This function was a logical test of whether the x >=50 condition was met and included return values for true and false elements of the test condition.
