---
title: "Assignment 3: Data Exploration"
author: Ricky Prophete, Section 4
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration. 

## Directions
1. Change "Student Name, Section #" on line 3 (above) with your name and section number.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "FirstLast_A03_DataExploration.Rmd") prior to submission.

The completed exercise is due on <>.


## Set up your R session

1. Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. **Be sure to add the `stringsAsFactors = TRUE` parameter to the function when reading in the CSV files.**
```{r,  message = FALSE}
#check working directory
getwd()
#setwd(/Users/ricpro/Documents/Duke MPP Coursework/Spring 2022/Environmental Data Analysis 872/Environmental_Data_Analytics_2022)
#load package
library(tidyverse)
#import datasets
Neonics <- read.csv("../Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE)
Litter <- read.csv("../Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE)

```

## Learn about your system

2. The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicologoy of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: The ecotoxicology of neonicotinoids may be relevant to understanding the effect of these insecticides on vulnerable insect populations -- with emphasis on species vital to agriclture (i.e. pollinators). These insecticides may also have adverse impacts on entire ecosystems by causing trophic cascades through the severe reduction of insect populations. 

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: Litter and woody debris can have positive impacts on ecosystems and biodiversity by serving as a source of nutrients and shelter for a number of species. Depending on the ecological context, the amount of litter can provide clues with respect to the overall health of an ecosystem. This debris can also serve as a source of wildfire fuel, and can be actively monitored to promote fire safety. 

4. How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here: 

> Answer: 
* Debris sampling is conducted at terrestrial NEON sites containing woody vegetation greater than 2m tall. 
* Ground traps are sampled once per year, with elevated trap sampling cadence varying accoring to the vegetation present on site. 
* Sampling occurs only in tower plots. Locations are selected at random.



## Obtain basic summaries of your data (Neonics)

5. What are the dimensions of the dataset?

```{r}
dim(Neonics) #4623 rows, 30 columns
```

6. Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?
```{r}
summary(Neonics$Effect)
```

> Answer: The most common effects are those on population and mortality. These are particularly relevant as they relate to the effect of neoniconinoids on a macro scale. Population effects and mortality can have downstream implications on ecosystem health.

7. Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r}
summary(Neonics$Species.Common.Name, 7)
```

> Answer: the most commonly studied species in this dataset are the Honey Bee, the Parasitc Wasp, the Buff Tailed Bumblebee, the Carniolan Honey Bee, the Bumble Bee, and the Italian Honeybee. These insects are relevant to human agriculture as all can serve as pollinators and wasps can serve in an additonal pest control capacity. Though wasps are generally carnivorous, they require high amounts of sugar and can inadvertently carry pollen between plants in search of nectar. 

8. Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r}
class(Neonics$Conc.1..Author.)
```

> Answer: Conc.1..Author is represented as a factor. This is due to the addition of the stringsAsFactors arguement which converted character vectors to factors when the dataset was imported into RStudio.


## Explore your data graphically (Neonics)

9. Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year. 
```{r}
freq_by_pub_year <- ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year))
#reproduce graph
freq_by_pub_year
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.
```{r}
freq_by_pub_year_color <- ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year, color = Test.Location))
#reproduce graph
freq_by_pub_year_color
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common test locations are labs and natural fields. Though labs and natural field genrally dominate after 1990 up to around 2020, their dominiance waxes and wanes over this period, with natural fields generally leading throughout the 1990s, and labs leading for the majority of 2010-2020.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

```{r}
bar_chart <- ggplot(Neonics, aes(x = Endpoint))+
  geom_bar()+
theme(axis.text.x = element_text(size=4))
#reproduce graph
bar_chart
```

> Answer: the two most common endpints are NOEL and LOEL. They are defined like so:
> NOEL = No-observable-effect-level: highest dose (concentration) producing effects not significantly different from responses of controls according to author's reported statistical test
> LOEL = Lowest-observable-effect-level: lowest dose (concentration) producing effects that were significantly different (as reported by authors) from responses of controls 

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018. 
```{r}
class(Litter$collectDate)
#convert from factor to date
Litter$collectDate <- as.Date(Litter$collectDate, "%Y-%m-%d")
#confirm conversion to date
class(Litter$collectDate)
#determine when litter was sampled in Aug 2018
unique(Litter$collectDate)
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
unique(Litter$plotID)
#compare with summary function
summary(Litter$plotID)
```

> Answer: There are 12 unique plots. The info obtained from 'unique' differs from 'summary' in that 'unique' provides a list of unique plotIDs whereas 'summary' provides a readout of the frequency of plotIDs

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
```{r}
bar_chart_Litter <- ggplot(Litter, aes(x = functionalGroup))+
  geom_bar()+
  theme(axis.text.x = element_text(size=6))
#reproduce graph
bar_chart_Litter
``` 

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup. 
```{r}
#boxplot
boxplot_Litter <- ggplot(Litter)+
  geom_boxplot(aes(x = dryMass, y = functionalGroup))
#reproduce boxplot
boxplot_Litter

#violin plot
violin_plot_Litter <- ggplot(Litter)+
  geom_violin(aes(x = dryMass, y = functionalGroup))
#reproduce violin plot
violin_plot_Litter
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: The boxplot is more effective as most observations for each functional group in this dataset are either relatively evenly distributed or are very close to the median with a few significant outliers. The chart scale therefore provides a much larger range than the majority of values in this dataset, making the relative distributions difficult to discern. 

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: Needles, mixed, and twigs/branches 