---
title: "Assignment 5: Data Visualization"
author: Ricky Prophete
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Monday, February 14 at 7:00 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse 
and cowplot packages. Upload the NTL-LTER processed data files for nutrients 
and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the [`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 
getwd()
library(tidyverse)
library(cowplot)
#load files
PeterPaul.ChemNutrients <- read.csv(
  "../Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv",
  stringsAsFactors = TRUE)
NEON.NIWO <- read.csv(
  "../Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)

#2
#check date format
class(PeterPaul.ChemNutrients$sampledate)
class(NEON.NIWO$collectDate)
#convert to date format
PeterPaul.ChemNutrients$sampledate <- as.Date(
  PeterPaul.ChemNutrients$sampledate, format = "%Y-%m-%d")
NEON.NIWO$collectDate <- as.Date(NEON.NIWO$collectDate, format = "%Y-%m-%d")

```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3
new_theme <- theme_bw(base_size = 13) + 
  theme(axis.text = element_text(color = "black"),
        legend.position = "bottom")
theme_set(new_theme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and `ylim()`).

```{r}
#4
#**install National Parks Color Palette**

#install.packages("devtools")
#devtools::install_github("katiejolly/nationalparkcolors")
library(nationalparkcolors)
names(park_palettes)

#create plot
plot_1 <- ggplot(PeterPaul.ChemNutrients, aes(x = po4, y = tp_ug)) +
  geom_point(aes(color = lakename))+
  geom_smooth(method = lm, color = "black")+
  scale_color_manual(values=park_palette("SmokyMountains"))+
  xlim(0, 50) +
  ylim(0, 150)
  print(plot_1)
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r}
#5
PeterPaul.ChemNutrients$month <- as.factor(PeterPaul.ChemNutrients$month)
#a
temp_plot <- ggplot(
  PeterPaul.ChemNutrients, aes(x = month, y = temperature_C, color = lakename)) +
  geom_boxplot()+
  theme(axis.title = element_text(size = 10))+
  scale_color_manual(values=park_palette("Arches"))
print(temp_plot)

#b
TP_plot <- ggplot(
  PeterPaul.ChemNutrients, aes(x = month, y = tp_ug, color = lakename)) +
  geom_boxplot()+
  theme(axis.title = element_text(size = 10))+
  scale_color_manual(values=park_palette("Arches"))
print(TP_plot)
  
#c
TN_plot <- ggplot(
  PeterPaul.ChemNutrients, aes(x = month, y = tn_ug, color = lakename)) +
  geom_boxplot()+
  theme(axis.title = element_text(size = 10))+
  scale_color_manual(values=park_palette("Arches"))
print(TN_plot)

#combined grid
temp_plot_b <- temp_plot + theme(legend.position = "none")
TP_plot_b <- TP_plot + theme(legend.position = "none") + ylim(0,80)
TN_plot_b <- TN_plot + theme(legend.position = "none") + ylim(0,1000)
legend <- get_legend(temp_plot + theme(legend.position = "right"))
plot_grid(temp_plot_b, TP_plot_b, TN_plot_b, 
          nrow = 4, ncol = 1, align = "hv",
          legend, labels = c("A","B","C"), label_size = 10, rel_heights = c(1, 1.25, 1.25), axis = "l")

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Median values for nutrient concentration experience increasing divergence over the summer months (Jun - Aug), with most of the variability attributable to Peter Lake. Median lake temperatures generally rise and fall together. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}
#6
Needles_plot <- ggplot(
  subset(NEON.NIWO, functionalGroup == "Needles"), aes(x = collectDate, y = dryMass, color = nlcdClass))+
  geom_point()+
  scale_color_manual(values=park_palette("Badlands"))+
  scale_x_date(date_breaks = "2 months", 
    date_labels = "%b %Y")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(Needles_plot)

#7
Needles_plot_2 <- ggplot(
  subset(NEON.NIWO, functionalGroup == "Needles"), 
                         aes(x = collectDate, y = dryMass))+
  geom_point()+
      facet_wrap(vars(nlcdClass))+
  scale_x_date(date_breaks = "5 months", 
               #5 months to improve legiblity on PDF
    date_labels = "%b %Y")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(Needles_plot_2)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: 7 is more effective as the facet view makes it much easier to interpret distributions of drymass observations within NLCD classes and compare observations across classes 
