---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: Ricky Prophete
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Monday, February 28 at 7:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
getwd()
library(tidyverse)
library(agricolae)

Peter.Paul.ChemNutrients <- read.csv(
  "../Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)
#Format dates
Peter.Paul.ChemNutrients$sampledate <- as.Date(
  Peter.Paul.ChemNutrients$sampledate, format = "%m/%d/%y")

#2
new_theme <- theme_bw(base_size = 13) + 
  theme(axis.text = element_text(color = "black"),
        legend.position = "bottom")
theme_set(new_theme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: The mean lake temp recorded during July does not change with depth
Ha: The mean lake temp recorded during July changes with depth 


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
Peter.Paul.Wrangled <- Peter.Paul.ChemNutrients %>% 
  select(lakename, year4, daynum, depth, temperature_C) %>%
  filter(daynum == 183:213) %>%
  drop_na()

#5
Plot1 <- ggplot(Peter.Paul.Wrangled, aes(x = depth, y = temperature_C))+
  geom_point()+
  ylim(0, 35)+
  labs(x = "Depth", y = "Temperature")+
  geom_smooth(method = "lm", color = "red", se = FALSE)
print(Plot1)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: Depth and temperature are inversely related. As depth increases, temperature decreases. The distribution of the points suggests that the inverse relationship between temperature and depth is most linear from around 0 to around 5 meters, after which temperature remains fairly constant despite further increases in depth. This inflection suggests the trend is not perfectly linear.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
Peter.Paul.Regression <- lm(data = Peter.Paul.Wrangled, temperature_C ~ depth)
summary(Peter.Paul.Regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: For every meter of incresed depth, water temperature is expected to decrease by 1.91 degrees celsius. 73% of the variability in temperature is explained by depth. This finding was computed with degress of freedom of 307, and is statistically different from zero.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
Peter.Paul.AIC <- lm(
  data = Peter.Paul.Wrangled, temperature_C ~ year4 + daynum + depth)

step(Peter.Paul.AIC)

#10 
Peter.Paul.Recommended <- lm(
  data = Peter.Paul.Wrangled, temperature_C ~ daynum + depth)
summary(Peter.Paul.Recommended)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of variables recommended by the AIC method are daynum and depth. This model explains 73.3% of the observed variance in temperature. This is a slight improvement over using only depth as an explanatory variable. However, compared to the effect of depth, we are less confident that the effect of daynum is statistically different from zero.

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
#categorical variable is lakename
#continuous variable is temperature

summary(Peter.Paul.Wrangled)
#ANOVA
Peter.Paul.Anova <- aov(data = Peter.Paul.Wrangled, temperature_C ~ lakename)
summary(Peter.Paul.Anova)

#Linear
Peter.Paul.Anova2 <- lm(data = Peter.Paul.Wrangled, temperature_C ~ lakename)
summary(Peter.Paul.Anova2)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: We do not reject the null hypothesis as there is no statistical difference in mean temperature across lakes. 1.6 percent of the variability in temperature is explained by the lake. This finding is computed with 300 degrees of freedom. 


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

Plot2 <- ggplot(Peter.Paul.Wrangled, aes(x = depth, y = temperature_C, color = lakename))+
  geom_point(alpha = 0.5)+
  #geom_smooth(method = "lm", se = FALSE)+
  geom_line(stat = "smooth", method = "lm", se = FALSE, alpha = 0.7)+ #using geom_line to make lines more legible as alpha doesn't work with geom_smooth
  ylim(0,35)+
  labs(x = "Depth", y = "Temperature")
print(Plot2)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(Peter.Paul.Anova)

#pairwise relationships
Lake.Groups <- HSD.test(Peter.Paul.Anova, "lakename", group = TRUE)
Lake.Groups

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Statistically speaking, all of the lakes have the same mean temp. No lake has a temperature that is statistically distinct from the other lakes.

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We could conduct a two sample t test to compare means between these two lakes.


