---
title: "Assignment 09: Data Scraping"
author: Ricky Prophete
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

# Total points:

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_09_Data_Scraping.Rmd") prior to submission.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1
getwd()
#load packages
library(tidyverse)
library(rvest)
library(lubridate)

#set theme
new_theme <- theme_bw(base_size = 13) + 
  theme(axis.text = element_text(color = "black"),
        legend.position = "bottom")
theme_set(new_theme)
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2019 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Change the date from 2020 to 2019 in the upper right corner.
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020>

Indicate this website as the as the URL to be scraped. (In other words, read the contents into an
`rvest` webpage object.)

```{r set.the.scraping.website}
#2
URL <- read_html(
  'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020')
URL
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Average Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to three separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values, with the first value being 36.0100.

```{r scrape.the.data}
#3
water.system.name <- URL %>% html_nodes('table:nth-child(7) tr:nth-child(1) td:nth-child(2)') %>% html_text()
pswid <- URL %>% html_nodes('td tr:nth-child(1) td:nth-child(5)') %>% html_text()
ownership <- URL %>% html_nodes('table:nth-child(7) tr:nth-child(2) td:nth-child(4)') %>% html_text()
max.withdrawals.mgd <- URL %>% html_nodes('th~ td+ td') %>% html_text()

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in order. You can overcome this by creating a month column in the same order the data are scraped: Jan, May, Sept, Feb, etc...

5. Plot the max daily withdrawals across the months for 2020

```{r create.a.dataframe.from.scraped.data}
#4
scraped.data.frame <- data.frame("Month" = c(1, 5, 9, 2, 6, 10, 3, 7, 11, 4, 8, 12),
                                 "Year" = rep(2020, 12),
                                 "Max_Withdrawals_mgd" = 
                                   as.numeric(max.withdrawals.mgd)) %>%
  arrange(Month) %>% 
  mutate(Water_System_Name = !!water.system.name,
       PSWID = !!pswid,
       Ownership = !!ownership,
       Date = my(paste(Month,"-",Year)))

#5
Plot1 <- ggplot(scraped.data.frame, aes(x = Date, y = Max_Withdrawals_mgd))+
  geom_line()+
  geom_smooth(method="loess",se=FALSE)+
  labs(title="Water Usage Data for Durham (2020)", y = "Max Daily Widthdrawals (mgd)", x = 'Date')+
  theme(plot.title = element_text(hjust = 0.5))
print(Plot1)


```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site scraped**.

```{r construct.a.scraping.function}
#6.
scraping.function <- function(the.year, PWSID){
  
  URL <- read_html(
    paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?', 'pwsid=', 
           PWSID, '&year=', the.year))
  
  print(URL)
  
  water.system.name.tag <- 'table:nth-child(7) tr:nth-child(1) td:nth-child(2)'
  pwsid.tag <- 'td tr:nth-child(1) td:nth-child(5)'
  ownership.tag <- 'table:nth-child(7) tr:nth-child(2) td:nth-child(4)'
  max.withdrawals.mgd.tag <- 'th~ td+ td'
  
  water.system.name <- URL %>% html_nodes(water.system.name.tag) %>% html_text()
  pwsid <- URL %>% html_nodes(pwsid.tag) %>% html_text()
  ownership <- URL %>% html_nodes(ownership.tag) %>% html_text()
  max.withdrawals.mgd <- URL %>% html_nodes(max.withdrawals.mgd.tag) %>% html_text()
  
  scraped.data.frame <- data.frame("Month" = c(1, 5, 9, 2, 6, 10, 3, 7, 11, 4, 8, 12),
                                   #"Month"= c('Jan', 'May', 'Sep', 'Feb', 'Jun', 'Oct', 'Mar', 'Jul', 'Nov', 'April', 'Aug', 'Dec'),
                                   "Year" = rep(the.year,12),
                                   "Max_Withdrawals_mgd" =
                                     as.numeric(max.withdrawals.mgd)) %>%
  arrange(Month) %>% 
  mutate(Water_System_Name = !!water.system.name,
       PWSID = !!pwsid,
       Ownership = !!ownership,
       Date = my(paste(Month,"-",Year)))
  
  return(scraped.data.frame)
}
```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7
df_1 <- scraping.function(2015,'03-32-010')
view(df_1)

Plot2 <- ggplot(df_1, aes(x = Date, y = Max_Withdrawals_mgd))+
  geom_line()+
  geom_smooth(method="loess",se=FALSE)+
  labs(title="Water Usage Data for Durham (2015)", y = "Max Daily Widthdrawals (mgd)", x = 'Date')+
  theme(plot.title = element_text(hjust = 0.5))
print(Plot2)

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares the Asheville to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8
df_2 <- scraping.function(2015,'01-11-010')
view(df_2)

#combine df_1 w/ df_2
combined_df <- bind_rows(df_1, df_2)

Plot3 <- ggplot(combined_df, aes(x = Date, y = Max_Withdrawals_mgd, color = Water_System_Name))+
  geom_smooth(method="loess",se=FALSE)+
  labs(title="Water Usage Data for Durham and Asheville (2015)", y = "Max Daily Widthdrawals (mgd)", x = 'Date')+
  theme(plot.title = element_text(hjust = 0.5))
print(Plot3)

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

```{r}
#9
the.years = rep(2010:2019)
the.location = '01-11-010'

df_3 <- lapply(X = the.years,
               FUN = scraping.function,
               PWSID = the.location)

df_3 <- map(the.years, scraping.function, PWSID = the.location)
view(df_3)

combined_df_2 <- bind_rows(df_3)

Plot4 <- ggplot(combined_df_2, aes(x = Date, y = Max_Withdrawals_mgd))+
  geom_line()+
  geom_smooth(method="loess",se=FALSE)+
  labs(title="Water Usage Data for Asheville", subtitle = "2010 - 2019", y = "Max Daily Widthdrawals (mgd)", x = 'Date')+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5))
plot(Plot4)

```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time? 

> Water usage appears to be increasing over time in Asheveille, with a marked inflection in use toward the end of the decade.